---
title: "Processing quantitative proteomics data with Features"
author:
- name: Laurent Gatto
package: Features
abstract: >
  This vignette describes how to process quantitative mass
  spectrometry data with Features: clearning up unneeded feature
  variables, adding an experimental design, filtering out contaminants
  and reverse hits, managing missing values, log-transforming,
  normalising and aggregating data.
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: Features.bib
vignette: >
  %\VignetteIndexEntry{Processing quantitative proteomics data with Features}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{Mass Spectrometry, MS, MSMS, Proteomics, Metabolomics, Infrastructure, Quantitative}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r env, message = FALSE, warning = FALSE, echo = FALSE}
library("Features")
```


# Introduction

## Reading data as `Features`

We are going to use a subset of the CPTAC study 6 containing
conditions A and B. The peptide-level data, as processed by MaxQuant
is available in the `msdata` package:


```{r msdata}
basename(f <- msdata::quant(pattern = "cptac", full.names = TRUE))
```

From the names of the columns, we see that the quantitative coumns,
starting with `"Intensity."` (note the space!) are at positions 56 to
61.

```{r cptac_cols}
names(read.delim(f))
(i <- grep("Intensity\\.", names(read.delim(f))))
```

We now read these data using the `readFeatures` function. The peptide
level expression data will be imported into R as an instance of class
`Features` namesd `cptac` with an assay named `peptides`. We also use
the `fnames` argument to set the rownames of the `peptides` assay to
the peptide sequences.

```{r read_cptac}
library("Features")
cptac <- readFeatures(f, ecol = i, sep = "\t", name = "peptides", fnames = "Sequence")
```

## Encoding the experimental design

Below we update the sample (column) annotations to encode the two
groups, 6A and 6B, and the original sample numbers.

```{r}
cptac$group <- rep(c("6A", "6B"), each = 3)
cptac$sample <- rep(7:9, 2)
colData(cptac)
```

## Filtering out contaminants and reverse hits


```{r}
filterFeatures(cptac, ~ Reverse == "")
```
```{r}
filterFeatures(cptac, ~ Potential.contaminant == "")
```

```{r}
library("magrittr")
cptac <- cptac %>%
    filterFeatures(~ Reverse == "") %>%
    filterFeatures(~ Potential.contaminant == "")
```

## Removinf up unneeded feature variables

The spreadsheet that was read above contained numerous variables that
are returned by MaxQuant, but not necessarily necessary in the frame
of a downstream statistical analysis. 

```{r}
rowDataNames(cptac)
```

The only ones that we will be needing below are the peptides sequences
and the protein identifiers. Below, we store these variables of
interest and filter them using the `selectRowData` function.

```{r}
rowvars <- c("Sequence", "Proteins", "Leading.razor.protein")
cptac <- selectRowData(cptac, rowvars)
rowDataNames(cptac)
```

## Managing missing values

Missing values can be very prominant in certain proteomics experiments
and need to be dealt with carefully. The first step is to assess their
presence accross samples and features. 


## TODO

- log-transformation
- normalisation
- aggregation



# Session information {-}

```{r sessioninfo, echo=FALSE}
sessionInfo()
```


# References {-}
