---
title: "Processing quantitative proteomics data with Features"
author:
- name: Laurent Gatto
package: Features
abstract: >
  This vignette describes how to process quantitative mass
  spectrometry data with Features: clearning up unneeded feature
  variables, adding an experimental design, filtering out contaminants
  and reverse hits, managing missing values, log-transforming,
  normalising and aggregating data.
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: Features.bib
vignette: >
  %\VignetteIndexEntry{Processing quantitative proteomics data with Features}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{Mass Spectrometry, MS, MSMS, Proteomics, Metabolomics, Infrastructure, Quantitative}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r env, message = FALSE, warning = FALSE, echo = FALSE}
library("Features")
```


# Introduction

## Reading data as `Features`

We are going to use a subset of the CPTAC study 6 containing
conditions A and B. The peptide-level data, as processed by MaxQuant
is available in the `msdata` package:


```{r msdata}
basename(f <- msdata::quant(pattern = "cptac", full.names = TRUE))
```

From the names of the columns, we see that the quantitative coumns,
starting with `"Intensity."` (note the space!) are at positions 56 to
61.

```{r cptac_cols}
names(read.delim(f))
(i <- grep("Intensity\\.", names(read.delim(f))))
```

We now read these data using the `readFeatures` function. The peptide
level expression data will be imported into R as an instance of class
`Features` namesd `cptac` with an assay named `peptides`. We also use
the `fnames` argument to set the rownames of the `peptides` assay to
the peptide sequences.

```{r read_cptac}
library("Features")
cptac <- readFeatures(f, ecol = i, sep = "\t", name = "peptides", fnames = "Sequence")
```

## Encoding the experimental design

Below we update the sample (column) annotations to encode the two
groups, 6A and 6B, and the original sample numbers.

```{r}
cptac$group <- rep(c("6A", "6B"), each = 3)
cptac$sample <- rep(7:9, 2)
colData(cptac)
```

## Filtering out contaminants and reverse hits


```{r}
filterFeatures(cptac, ~ Reverse == "")
```
```{r}
filterFeatures(cptac, ~ Potential.contaminant == "")
```

```{r}
library("magrittr")
cptac <- cptac %>%
    filterFeatures(~ Reverse == "") %>%
    filterFeatures(~ Potential.contaminant == "")
```

## Removinf up unneeded feature variables

The spreadsheet that was read above contained numerous variables that
are returned by MaxQuant, but not necessarily necessary in the frame
of a downstream statistical analysis. 

```{r}
rowDataNames(cptac)
```

The only ones that we will be needing below are the peptides sequences
and the protein identifiers. Below, we store these variables of
interest and filter them using the `selectRowData` function.

```{r}
rowvars <- c("Sequence", "Proteins", "Leading.razor.protein")
cptac <- selectRowData(cptac, rowvars)
rowDataNames(cptac)
```

## Managing missing values

Missing values can be very prominant in certain proteomics experiments
and need to be dealt with carefully. The first step is to assess their
presence accross samples and features. But before being able to do so,
we need to replace 0 by `NA`, given that MaxQuant encodes missing data
with a 0 using the `zeroIsNA` function.

```{r}
cptac <- zeroIsNA(cptac)
nNA(cptac)
```

The output of the `nNA` function tells us that 

- there are currently close to 50% is missing values in the data;
- there are 4051 peptides with 0 missing values, 989 with a single
  missing values, ... and 3014 peptides composed of only missing
  values;
- the range of missing values in the 6 samples is comparable and
  ranges between 4651 and 5470.

In this dataset, we have such a high number of peptides without any
data because the 6 samples are a subset of a larger dataset, and these
peptides happened to be absent in groups A and B. Below, we use
`filterNA` to remove these peptides. By setting `pNA` to 0.9, we allow
petides that have up to 90% of missing values, inclusing in this case
those that have only one value (i.e 5/6 percent of missing values).

```{r}
cptac <- filterNA(cptac, pNA = .9)
```

If we wanted to remove all peptides that have one or more missing
values, we could simply use the default value of `pNA = 0`.

## Data transformation

When analysing continuous data using parametric methods (such as
t-test or linear models), it is often necessary to log-transform the
data. The figure below (left) show that how our data is mainly composed of
small values with a long tail of larger ones, which is a typical
pattern of quantitative omics data.

Below, we use the `logTransform` function to log2-transform our
data. This time, instead of overwriting the peptides assay, we are
going to create a new one to contain the log2-transfromed data.

```{r}
cptac <- addAssay(cptac, logTransform(cptac[[1]]), name = "cptac_log")
cptac
```

```{r, fig.cap = "Quantitative data in its original scale (left) and log2-transformed."}
par(mfrow = c(1, 2))
limma::plotDensities(assay(cptac[[1]]))
limma::plotDensities(assay(cptac[[2]]))
```

## TODO

- normalisation
- aggregation
- visualisation

# Session information {-}

```{r sessioninfo, echo=FALSE}
sessionInfo()
```


# References {-}
