---
title: "Supported input formats for `readQFeatures()`"
author:
  - name: "Karolína Kryštofová"
  - name: Laurent Gatto
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('QFeatures')`"
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show

vignette: >
  %\VignetteIndexEntry{readQFeatures() input format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
link-citations: true
link-bibliography: true
---

## Methods

### Datasets

This vignette demonstrates the use of the *QFeatures* package's
`readQFeatures()` function to import data produced by popular
third-party software. For this purpose, subsets of the following
previously publicly available datasets have been used:

+---------------------------+-----------+------+---------+------------+
| Citation                  | PXD ID    | Mode | Label   | Code       |
+===========================+===========+======+=========+============+
| Van Puyvelde et al., 2022 | PXD028735 | DIA  | LFQ     | A_DIA_LFQ  |
|                           |           +------+---------+------------+
|                           |           | DDA  | LFQ     | A_DDA_LFQ  |
+---------------------------+-----------+------+---------+------------+
| Derks et al., 2022        | PXD029531 | DIA  | plexDIA | B_DIA_plex |
+---------------------------+-----------+------+---------+------------+
| Christoforou et al., 2016 | PXD001279 | DDA  | TMT     | C_DDA_TMT  |
+---------------------------+-----------+------+---------+------------+

Specifically, these subsets consist of these files:

+------------+-------------------------------------------------------+
| Code       | Original raw file name                                |
+------------+-------------------------------------------------------+
| A_DIA_LFQ  | LFQ_timsTOFPro_diaPASEF_Condition_A_Sample_Alpha_01.d |
|            +-------------------------------------------------------+
|            | LFQ_timsTOFPro_diaPASEF_Condition_A_Sample_Alpha_02.d |
|            +-------------------------------------------------------+
|            | LFQ_timsTOFPro_diaPASEF_Condition_A_Sample_Alpha_03.d |
|            +-------------------------------------------------------+
|            | LFQ_timsTOFPro_diaPASEF_Condition_B_Sample_Alpha_01.d |
|            +-------------------------------------------------------+
|            | LFQ_timsTOFPro_diaPASEF_Condition_B_Sample_Alpha_02.d |
|            +-------------------------------------------------------+
|            | LFQ_timsTOFPro_diaPASEF_Condition_B_Sample_Alpha_03.d |
+------------+-------------------------------------------------------+
| A_DDA_LFQ  | LFQ_Orbitrap_DDA_Condition_A_Sample_Alpha_01.raw      |
|            +-------------------------------------------------------+
|            | LFQ_Orbitrap_DDA_Condition_A_Sample_Alpha_02.raw      |
|            +-------------------------------------------------------+
|            | LFQ_Orbitrap_DDA_Condition_A_Sample_Alpha_03.raw      |
|            +-------------------------------------------------------+
|            | LFQ_Orbitrap_DDA_Condition_B_Sample_Alpha_01.raw      |
|            +-------------------------------------------------------+
|            | LFQ_Orbitrap_DDA_Condition_B_Sample_Alpha_02.raw      |
|            +-------------------------------------------------------+
|            | LFQ_Orbitrap_DDA_Condition_B_Sample_Alpha_03.raw      |
+------------+-------------------------------------------------------+
| B_DIA_plex | wJD803.raw                                            |
|            +-------------------------------------------------------+
|            | wJD804.raw                                            |
|            +-------------------------------------------------------+
|            | wJD815.raw                                            |
+------------+-------------------------------------------------------+
| C_DDA_TMT  | Replicate1_fraction1.raw                              |
|            +-------------------------------------------------------+
|            | Replicate1_fraction2.raw                              |
+------------+-------------------------------------------------------+

The data files are available in the `MsDataHub` package (

```{r MsDataHub}
library("MsDataHub")
MsDataHub() |>
    dplyr::filter(grepl("19137577", SourceUrl)) |>
    dplyr::select(Title, Description)
```

Identifications and quantifications performed on these datasets were
used in combination with the following software:

```{r dataset_matrix, echo=FALSE}
suppressPackageStartupMessages(library(ComplexHeatmap))

# All possible softwares, labels and report levels
sws <- c('MaxQuant', 'DIA-NN', 'Spectronaut', 'Proteome Discoverer', 'sage', 'FragPipe', 'PEAKS')
label <- c('LFQ', 'multiplex')

# Create a combination dataset, by default leave value cells blank
## value: 0 for not yet processed data, 1 for already processed data
## dataset: what dataset will be used
d <- expand.grid(software = sws, label = label)
d$value <- 'no'
d$dataset <- ''

# Fill in manually with currently known information
d$dataset[which(d$label == 'LFQ' & d$software == 'Spectronaut')] <- 'A_DIA_LFQ'
d$dataset[which(d$label == 'LFQ' & d$software == 'DIA-NN')] <- 'A_DIA_LFQ'
d$dataset[which(d$label == 'LFQ' & d$dataset == '')] <- 'A_DDA_LFQ'
d$dataset[which(d$label == 'multiplex' & d$software %in% c('DIA-NN', 'Spectronaut'))] <- 'B_DIA_plex'
d$dataset[which(d$dataset == '')] <- 'C_DDA_TMT'

d$value[which(d$software %in% c('DIA-NN', 'MaxQuant', 'sage', 'FragPipe'))] <- 'yes'
d$value[which(d$software == 'Spectronaut' & d$label == 'LFQ')] <- 'yes'
d$value[which(d$software == 'FragPipe')] <- 'yes'
d$value[which(d$software == 'PEAKS' & d$label == 'LFQ')] <- 'yes'

# Dataset to matrix for plotting
d <- tidyr::unite(d, col, label)
#lvls <- c('LFQ_PSM', 'LFQ_peptide', 'LFQ_PG',
#          'multiplex_PSM', 'multiplex_peptide', 'multiplex_PG')
#d$col <- factor(d$col, levels = lvls)
#d <- dplyr::arrange(d, col)

m <- d |> dplyr::select(-dataset) |> tidyr::spread(key = 'col', value = 'value')
rownames(m) <- m$software
m$software <- NULL
m <- as.matrix(m)
colnames(m) <- gsub('LFQ_', '', fixed = TRUE, colnames(m))
colnames(m) <- gsub('multiplex_', '', fixed = TRUE, colnames(m))

# Create a helped matrix storing the text annotations
m_anno <- d |> dplyr::select(-value) |> tidyr::spread(key = 'col', value = 'dataset')
rownames(m_anno) <- m_anno$software
m_anno$software <- NULL
m_anno <- as.matrix(m_anno)
colnames(m_anno) <- gsub('LFQ_', '', fixed = TRUE, colnames(m_anno))
colnames(m_anno) <- gsub('multiplex_', '', fixed = TRUE, colnames(m_anno))

# Create a plot using the ComplexHeatmap package
colors = structure(c('#8BB135', '#FAA613', 'grey95'),
                   names = c("yes", "TBD", "no"))

#top_ha <- HeatmapAnnotation(foo = anno_block(labels = c("LFQ", "multiplex"),
#                                             labels_gp = gpar(fontface = 2),
#                                             gp = gpar(col = 'white')))

Heatmap(m, name = 'Processed?',
        col = colors,
        cluster_rows = FALSE, cluster_columns = FALSE,
        row_names_side = "left",
        column_names_side = "top",
        column_names_centered = TRUE,
        column_names_rot = 0,
        #column_split = rep(1:2, each = 3), column_title = NULL,
        border = TRUE,
#        top_annotation = top_ha,
        cell_fun = function(j, i, x, y, width, height, fill) {
          grid.text(m_anno[i, j], x, y, gp = gpar(fontsize = 10))
        })
```

### Existing search outputs

Example outputs for LFQ quantification using DIA-NN, Spectronaut and
PEAKS were sourced from the ProteoBench website. IDs of these outputs
on the ProteoBench website are as follows:

-   DIA-NN:
    -   DIA-NN_20250714_074145 (.parquet output)
    -   DIA-NN_20250606_103313 (.tsv output)
-   Spectronaut:
    -   Spectronaut_20250609_135453
-   PEAKS:
    -   PEAKS_20250714_150458

As an example output for multiplex quantification using DIA-NN, a
search result of the plexDIA dataset was sourced from the MassIVE
repository (MSV000088302).

### New search outputs

Following software versions were used to produce search results used
in this vignette:

-   MaxQuant 2.7.0
-   sage 0.14.7
-   FragPipe 23.1

## Introduction

Below are described individual outputs and their processing using the
`readQFeatures()` functions. When applicable, we demonstrate how to
read data on PSM, precursor, as well as protein group level.

For general explanation of the `QFeatures` class and detailed
description of individual arguments taken by the `readQFeatures()`
group of functions, consult the `readQFeatures()` manual page of this
[vignette](https://rformassspectrometry.github.io/QFeatures/articles/readQFeatures.html).

To initiate the session, we will load the `QFeatures` package.

```{r library, message = FALSE}
library(QFeatures)
```
For a future reference, we will also save parts of the output filenames for individual datasets into variables:

```{r}
# Dataset A
prefixAdda <- './data/vanPuyvelde_2022_LFQ_DDA_'
prefixAdia <- './data/vanPuyvelde_2022_LFQ_DIA_'

# Dataset B
prefixB <- './data/Derks_2022_plex_DIA_'

# Dataset C
prefixC <- './data/Christoforou_2016_TMT_DDA_'
```


## MaxQuant

MaxQuant produces several output .txt files. In order to obtain information from several levels of the search, we can look at the *evidence.txt*, *peptides.txt* and *proteinGroups.txt* files.

Here we will process the results of a multi-set label-free experiment. First we will read the *evidence.txt* file storing information about PSM-level data:

```{r mq-lfq-psm, message=FALSE}
# Read evidence.txt as R dataframe
fnMaxquantLFQevidence <- paste0(prefixAdda, 'MaxQuant_evidence.txt')
dataMaxquantLFQevidence <- read.delim(fnMaxquantLFQevidence)

nrow(dataMaxquantLFQevidence)

# Read as QFeatures object using Intensity as quantitative column
qfMaxquant <- readQFeatures(dataMaxquantLFQevidence, quantCols = 'Intensity',
                            runCol = 'Experiment')
names(qfMaxquant) <- paste('psm', names(qfMaxquant), sep = '_')

qfMaxquant
```

Next we will read the peptide-level results from a *peptides.txt* file and append this to the `QFeatures` object as a new assay:

```{r mq-lfq-peptide, message=FALSE}
# Read peptides.txt file as a R dataframe
fnMaxquantLFQpeptide <- paste0(prefixAdda, 'MaxQuant_peptides.txt')
dataMaxquantLFQpeptide <- read.delim(fnMaxquantLFQpeptide)

nrow(dataMaxquantLFQpeptide)

# Get indices of intensity columns
i <- grep('Intensity.', colnames(dataMaxquantLFQpeptide), fixed = TRUE)

qfMaxquant <- addAssay(qfMaxquant,
                       readQFeatures(dataMaxquantLFQpeptide, quantCols = i, fnames = 'Sequence')[[1]],
                       name = 'peptides')
qfMaxquant
```

We see that a new assay has been appended to `QFeatures` object. Finally, we will append the protein group-level in the same manner. Here we will use the *LFQ.intensity* columns:

```{r mq-lfq-pg, message=FALSE}
# Read peptides.txt file as a R dataframe
fnMaxquantLFQprotein <- paste0(prefixAdda, 'MaxQuant_proteinGroups.txt')
dataMaxquantLFQprotein <- read.delim(fnMaxquantLFQprotein)

nrow(dataMaxquantLFQprotein)

# Get indices of LFQ intensity columns
i <- grep('LFQ.intensity.', colnames(dataMaxquantLFQprotein), fixed = TRUE)

qfMaxquant <- addAssay(qfMaxquant,
                       readQFeatures(dataMaxquantLFQprotein, quantCols = i, fnames = 'Protein.IDs')[[1]],
                       name = 'proteinGroups')
qfMaxquant
```

In the next section, we will demonstrate how to read data from a TMT-labeled experiment consisting of two runs:

```{r mq-tmt, message=FALSE}

# Read the txt output
fnMaxquantTMTevidence <- paste0(prefixC, 'MaxQuant_evidence.txt')
dataMaxquantTMTevidence <- read.delim(fnMaxquantTMTevidence)

# Get the reporter intensity column IDs
i <- grep('Reporter.intensity.\\d+', colnames(dataMaxquantTMTevidence))

# Create a QFeatures object
qfMaxquantTMT <- readQFeatures(dataMaxquantTMTevidence,
                               quantCols = i,
                               runCol = 'Raw.file',
                               fnames = 'Sequence')
qfMaxquantTMT
```
Here we see that a separate experiment has been created for each run with 10 columns corresponding to the 10 TMT channels.

```{r mq-tmt-2, message=FALSE}
head(assay(qfMaxquantTMT))
```

## DIA-NN

DIA-NN versions 1.9.0 and below produce a main *.tsv* search result file, which has been from version 2.0.0 up solely replaced by a *.parquet* file.

DIA-NN *.tsv* reports can be read using the `readQFeaturesFromDIANN()` function:

```{r diann-tsv, message = FALSE}
# Load DIA-NN .tsv report
fnDiannLFQ <- paste0(prefixAdia, 'DIANN_report.tsv')

dataDiannLFQ <- read.delim(fnDiannLFQ)
qfDiannLFQ <- readQFeaturesFromDIANN(dataDiannLFQ, runCol = 'Run')

qfDiannLFQ
```

In order to read a *.parquet* file in R, we need to use the `arrow` package providing interface to the Arrow C++ library. After reading this file however, we can work with the resulting dataframe in the same manner as we are used to in case of the *.tsv* report.

```{r diann-parquet, message=FALSE}
# Get path to the input .parquet file
fnDiannLFQparquet <- paste0(prefixAdia, 'DIANN_report.parquet')

# Read .parquet file
dataDiannLFQparquet <- arrow::read_parquet(fnDiannLFQparquet)
qfDiannParquet <- readQFeaturesFromDIANN(dataDiannLFQparquet, runCol = 'Run')

qfDiannParquet
```

As you can see, both experiments consist of the same run names as both searches have been performed on the same set of raw data. The numbers of rows in each `SummarizedExperiment` however differ between those two reports, as both searches have been performed using different software versions, as well as different search parameters.

To correctly parse a plexDIA experiment, it is necessary to set the `multiplexing` parameter to `"mTRAQ"`:

```{r diann-plexdia, message=FALSE}
# Get path to the input plexDIA search result
fnDiannPlex <- paste0(prefixB, 'DIANN_report.tsv')

dataDiannPlex <- read.delim(fnDiannPlex)
qfDiannPlex <- readQFeaturesFromDIANN(dataDiannPlex, runCol = 'Run',
                                      multiplexing = 'mTRAQ')

qfDiannPlex
```

The run names in this output file are not the most informative with regards to the samples. We will now edit the sample metadata to contain more meaningful sample annotation.

```{r}
colData(qfDiannPlex)
rownames(colData(qfDiannPlex))

# All runs were performed on the same sample, in 3 technical replicates:
qfDiannPlex$sample <- 'mixed standard'
qfDiannPlex$rep <- rep(1:3, each = 3)
qfDiannPlex$label <- paste0('mTraq d', rep(c(0, 4, 8), times = 3))

colData(qfDiannPlex)
```

## sage

The **sage** search engine stores quantification data either in the *lfq.tsv* or *tmt.tsv* file based on the quantification used.

As for label-free quantification, the *lfq.tsv* file contains estimated quantities of identified proteoforms and is grouped on modified sequence level:

```{r sage-lfq, message = FALSE, warning=FALSE}
fnSageLFQ <- paste0(prefixAdda, 'sage_lfq.tsv')
dataSageLFQ <- read.delim(fnSageLFQ)

# Get indices of intensity columns
intensityCol <- grep('.mzML', colnames(dataSageLFQ), fixed = TRUE)

qfSageLFQ <- readQFeatures(dataSageLFQ, quantCols = intensityCol, name = 'peptides')
qfSageLFQ
```

As for TMT-based quantification, the PSM-level quantification is included in the *tmt.tsv* file.

```{r sage-tmt, message = FALSE, warning = FALSE}
# First, we will load the tmt.tsv file
fnSageTMT <- paste0(prefixC, 'sage_tmt.tsv')
dataSageTMT <- read.delim(fnSageTMT)
```

Upon inspection, we can see that peptide identification information is missing in this file:

```{r}
colnames(dataSageTMT)
```

We can source this from the main results.sage.tsv file and append this information to the quantification data frame:

```{r sage-tmt-2, message = FALSE, warning = FALSE}
fnSageTMTident <- paste0(prefixC, 'sage_results.sage.tsv')
dataSageTMTident <- read.delim(fnSageTMTident)

dataSageTMTfinal <- merge(dataSageTMT, dataSageTMTident, by = c('filename', 'scannr'))

# We also extract the indices of the quantification columns
iQuantSageTMT <- grep('tmt_', colnames(dataSageTMTfinal), fixed = TRUE)

# Finally we can load this dataframe using the readQFeatures() function
qfSageTMT <- readQFeatures(dataSageTMTfinal, quantCols = iQuantSageTMT)
qfSageTMT
```

A more straightforward way is to use the `sager` library to quickly load TMT quantification data from both *.tsv* output files into a `QFeatures` object.

```{r sager, message = FALSE}
qfSager <- sager::sageQFeatures(fnSageTMT, fnSageTMTident)
qfSager
```

## FragPipe

**FragPipe** produces several outputs. Following block shows the processing of a label-free multi-set experiment. First we can load the *psm.tsv* files that are produced separately for each sample-biological replicate combination as specified during FragPipe configuration. In our case, there has been a separate file created for each run:

```{r fragpipe-lfq, message = FALSE}
# Get the path to the directory containing FragPipe outputs
folderFpipe <- './data/'

# Get all paths to psm.tsv files
fnFpipeLFQpsm <- list.files(path = folderFpipe, pattern = 'FragPipe_\\w_\\d_psm\\.tsv')

# Within following 'for loop', load each file and append to a QFeatures object
for (fn in fnFpipeLFQpsm) {
  dataFpipeLFQpsm <- read.delim(paste0(folderFpipe, fn))

  if(which(fnFpipeLFQpsm == fn) == 1) { # with the first file, create a new QFeatures object
    qfFpipeLFQ <- readQFeatures(dataFpipeLFQpsm, quantCols = 'Intensity', name = fn)
  } else { # with other files, just append to an existing QFeatures object
    qfFpipeLFQ <- addAssay(qfFpipeLFQ,
                           readQFeatures(dataFpipeLFQpsm, quantCols = 'Intensity')[[1]],
                           name = fn)
  }
}

qfFpipeLFQ
```
The names of the assays are based on the (rather long) filenames, they were derived from. We can shorten these:

```{r fragpipe-lfq-names, message=FALSE}
names(qfFpipeLFQ) <- sub('vanPuyvelde_2022_LFQ_DDA_FragPipe_(\\w_\\d_psm)\\.tsv', '\\1', names(qfFpipeLFQ))
qfFpipeLFQ
```
The processing of peptide and protein-level outputs is analogical to MaxQuant processing.

In the following section, we demonstrate the processing of TMT-labelled multi-set experiment. It consists of two runs named *Fraction1* and *Fraction2*. Just like in the case of a label-free experiment, there is a separate *psm.tsv* file produced for each run:

```{r fragpipe-tmt, message=FALSE}
fnFpipeTMTpsm <- list.files(path = folderFpipe, pattern = 'FragPipe_Fraction\\d_psm\\.tsv')

for (fn in fnFpipeTMTpsm) {
  dataFpipeTMTpsm <- read.delim(paste0(folderFpipe, fn))
  intensityColFpipe <- grep('Intensity\\.', colnames(dataFpipeTMTpsm))

  if(which(fnFpipeTMTpsm == fn) == 1) { # with the first file, create a new QFeatures object
    qfFpipeTMT <- readQFeatures(dataFpipeTMTpsm, quantCols = intensityColFpipe, name = fn)


  } else { # with other files, just append to an existing QFeatures object
    qfFpipeTMT <- addAssay(qfFpipeTMT,
                           readQFeatures(dataFpipeTMTpsm, quantCols = intensityColFpipe)[[1]],
                           name = fn)
  }
}

qfFpipeTMT
```
